Multilevel computation of glacier geometry \\ from Stokes dynamics

### 2021-06-05 paper/msg.tex
Edits on IDO material, including change of sign.

\newcommand{\pp}{{\text{p}}}
\newcommand{\qq}{{\text{q}}}
\newcommand{\rr}{{\text{r}}}

@@

We model the ice as an incompressible, very-viscous \cite{Acheson1990}, non-Newtonian fluid subject to Glen's shear-thinning flow law \cite{GreveBlatter2009}; see also \cite[Chapter 1]{FowlerNg2021}.  Allowing any Glen exponent $n\ge 1$, the equations inside $\Lambda_s$ are
\begin{align}
- \nabla \cdot \tau + \nabla p &= \rhoi \bg &&\text{\emph{stress balance}} \label{eq:forcebalance} \\
\nabla \cdot \bu &= 0 &&\text{\emph{incompressibility}} \label{eq:incompressible} \\
\tau &= B_n |D\bu|^{(1/n) - 1} D\bu  &&\text{\emph{flow law}} \label{eq:viscflowlaw}
\end{align}
The solution fields here are the velocity $\bu$, pressure $p$, and deviatoric stress $\tau$.  Equations \eqref{eq:forcebalance} and \eqref{eq:viscflowlaw} form the momentum conservation (stress balance) model, while \eqref{eq:incompressible} follows from constant density and mass conservation.

@@

In computations we will use constants $n=3$, ice density $\rhoi=910 \,\text{kg}\,\text{m}^{-3}$ \cite{Huybrechtsetal1996}, and gravity $\bg=\left<0,0,-g\right>$, with $g=9.81\,\text{m}\,\text{s}^{-2}$.  The ice hardness $B_n$ is a constant because we assume isothermal conditions \cite{GreveBlatter2009}, with the value $B_3=6.8082\times 10^7\,\text{Pa}\,\text{s}^{1/3}$ \cite{Huybrechtsetal1996} used in computations.

For the $n=1$ (Newtonian) Stokes equations one would write \eqref{eq:viscflowlaw} as $\tau = 2\nu D\bu$ with viscosity $\nu>0$, but exponents $n>1$ imply an effective viscosity function of $|D\bu|$.  This would be singular in the limit of small strain rates, and so, motivated by the actual finite viscosity of glacier ice \cite{GreveBlatter2009}, we define the regularized effective viscosity using $\eps>0$,
\begin{equation}
\nu_\eps = \frac{1}{2} B_n \left(|D\bu|^2 + \eps\, D_0^2\right)^{(\pp-2)/2}, \label{eq:regeffvisc}
\end{equation}
where $\pp=(1/n)+1$, with $\pp=4/3$ in computations.  The constant $D_0$ defines a strain-rate scale for glacier flow; values $D_0 = 1 \,\text{a}^{-1}$ and $\eps = 10^{-4}$ are used in computations.

@@

We may now state the strong form of the SIGP, as follows, by including all of the above conditions, and also by eliminating $\tau$ from \eqref{eq:forcebalance} and \eqref{eq:viscflowlaw}:
\begin{align}
s - b &\ge 0 && \text{on $\Omega$} \label{eq:strongform} \\
- \bu|_s \cdot \bn_s - a &\ge 0 && \text{\emph{same}} \notag \\
(s - b) (- \bu|_s \cdot \bn_s - a) &= 0 && \text{\emph{same}} \notag \\
- \nabla \cdot \left(2 \nu_\eps\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{on $\Lambda_s$} \notag \\
\nabla \cdot \bu &= 0 && \text{\emph{same}} \notag \\
\left(2 \nu_\eps D\bu - pI\right) \bn &= \bzero && \text{on $\partial \Lambda_s \setminus \Gamma_0$} \notag \\
\bu &= \bzero && \text{on $\Gamma_0$} \notag
\end{align}
(Definition \eqref{eq:regeffvisc} of $\nu_\eps$ is also needed.)  The first three statements in \eqref{eq:strongform} form the NCP, but this problem is coupled to the boundary value problem formed by the last four statements.

@@

Denote by $W^{k,r}$ the Sobolev space \cite{Evans2010} of functions with $k$ derivatives which are $r$th-power integrable.  Suppose for the moment that $\Lambda \subset \RR^{d+1}$ is a fixed, bounded domain on which we prescribe a Dirchlet condition $\bu=0$ on some positive-measure portion of the boundary, and otherwise suppose the boundary is stress-free (Neumann boundary).  Let $\pp=(1/n)+1$ as in \eqref{eq:regeffvisc}, and $\qq=(1-\pp^{-1})^{-1}=n+1$ the conjugate exponent; $\pp=4/3$ and $\qq = 4$ if $n=3$.  Let $W_0^{1,\pp}(\Lambda)^{d+1}$ be the space of velocity functions, zero along the Dirichlet boundary, and let
\begin{equation}
\mathcal{M}_{\text{GS}} = W_0^{1,\pp}(\Lambda)^{d+1} \times L^\qq(\Lambda)  \label{eq:mixed}
\end{equation}
be the (mixed) space of admissible velocity and pressure pairs.  The Glen-Stokes solution $(\bu,p) \in \mathcal{M}_{\text{GS}}$ should satisfy the weak form
\begin{equation}
F_{\text{GS}}(\bu,p)[\bv,q] = \int_\Lambda 2 \nu_\eps D\bu : D\bv - p \Div\bv - (\Div\bu) q - \rhoi \bg \cdot \bv\,d\bx = 0 \label{eq:glenstokesweak}
\end{equation}
for all $(\bv,q) \in \mathcal{M}_{\text{GS}}$.

@@

However, we must go beyond the problem on predetermined ice geometry to solve the weak form of \eqref{eq:strongform}.  Again suppose $\Omega \subset \RR^d$ is a fixed map-plane domain with $C^1$ boundary, and that the bed elevation $b$ is in $W^{1,\qq}(\Omega)$.  The Sobolev space $W^{1,\qq}(\Omega)$ supports a linear trace operator which defines the symbol $|_{\partial \Omega}$ \cite[Section 5.5]{Evans2010}.  Let
\begin{equation}
\mathcal{K} = \{s \in W^{1,\qq}(\Omega) \,:\, s \ge b \, \text{ and } \, s\big|_{\partial\Omega} = b\big|_{\partial\Omega}\}  \label{eq:Kconstraintset}
\end{equation}
be the closed and convex set of admissible surface elevations of an isolated glacier or ice sheet.  From now on we also assume that $n>1$.  Because $\qq = n+1 > d$, each $s \in W^{1,\qq}(\Omega)$ is continuous \cite[Morrey's inequality, section 5.6.2]{Evans2010}, and so there are no cliffs in the $s$ geometry.

However, it is not clear that the correct Sobolev space has been identified in \eqref{eq:Kconstraintset}, and in fact we must further suppose that
\begin{equation}
s\in \mathcal{K} \text{ defines an icy domain } \Lambda_s \text{ which has a $C^1$ upper surface.} \label{eq:quixotic}
\end{equation}
Quixotic assumption \eqref{eq:quixotic} would follow from a sufficiently-strong regularity result for the theory of the current paper, which we cannot offer.  However, the existing theory of well-posedness for the SIA model includes a version of definition \eqref{eq:Kconstraintset} \cite{JouvetBueler2012}, wherein a power of the ice thickness, $u=H^{2q/(q-1)}$, is in $\mathcal{K} = \{u \ge 0\} \subset W^{1,\qq}(\Omega)$, and furthermore in the SIA the surface $s$ solves an elliptic equation and is thus smooth on the set $\{s<b\} \subset \Omega$.

We may now define the ice dynamics operator (IDO), a map which uses the trace of the velocity solution to the Glen-Stokes problem above (Figure \ref{fig:idoaction}), as follows.  For each surface elevation function $s \in \mathcal{K}$ the IDO result $\Phi(s)$ is the normal component of the upper-surface trace of the unique velocity solution $\bu$ of \eqref{eq:glenstokesweak} over the icy domain $\Lambda_s$, but extended by zero to all of $\Omega$:
\begin{equation}
\Phi(s) = \begin{cases} \bu|_s \cdot \bn_s, & s > b, \\
                        0, & \text{otherwise}. \end{cases} \label{eq:ido}
\end{equation}
Here $\bu|_s$ denotes a trace operator on $W^{1,\pp}(\Lambda_s)^{d+1}$; compare \eqref{eq:mixed}.  Note that the output $\Phi(s)$, a scalar function on $\Omega$, may jump or have other singularities at the glacier margin, but we will only need $\Phi(s)$ to be in the dual of $W^{1,\qq}(\Omega)$, permitting substantial irregularities.

\begin{figure}[t]
\begin{center}
\includegraphics[width=\textwidth]{genfigs/idoaction.pdf}
\end{center}
\caption{The ice dynamics operator $\Phi$ maps the surface elevation $s$ (dashed) to the normal ice surface motion $\Phi(s)=\bu|_s \cdot \bn_s$, equation \eqref{eq:ido}.  Both the input and output of $\Phi$ are defined on all of $\Omega$.}
\label{fig:idoaction}
\end{figure}

FIXME TALK ABOUT IDO.  1 FORM IN SIA CASE. 2 LINEARIZATION DENSE.  3 SPECTRUM OF IDO DETERMINES STABILITY OF EXPLICIT TIME-STEPPING. 4 DIAGONAL OF IDO UNKNOWN BUT POSITIVE?

Using the IDO, and recalling that the CMB $a$ is defined on all of $\Omega$, we may immediately re-write the \eqref{eq:strongform} in a simpler strong form, namely an NCP with a hidden dynamical problem:
\begin{align}
s - b &\ge 0  \label{eq:idostrongform} \\
- \Phi(s) - a &\ge 0 \notag \\
(s - b) (- \Phi(s) - a) &= 0 \notag
\end{align}
Here the statements hold on all of $\Omega$ and the data $a,b$ are clear.  The solution is the ice surface $z=s(x,y)$, but $\bu,p$ are hidden variables within the evaluation of $\Phi(s)$.

For $s \in \mathcal{K}$ and $r \in W^{1,\qq}(\Omega)$ we next define a functional which is nonlinear in $s$:
\begin{equation}
F(s)[r] = - \ip{\Phi(s)}{r} = - \int_\Omega \Phi(s)\, r \,dx dy. \label{eq:sigpfunctional}
\end{equation}
One computes $F(s)[r]$ by defining the icy domain $\Lambda_s$ as in \eqref{eq:lambdas}, then solving the (weak form) Glen-Stokes problem \eqref{eq:glenstokesweak}, then evaluating the normal component of the surface trace of the velocity \eqref{eq:ido} to compute the IDO $\Phi(s)$, and then integrating the result against the test function $r$.  Implicit in \eqref{eq:sigpfunctional} is that $\Phi(s)$ is in the dual space of $W^{1,\qq}(\Omega)$, so it possesses minimal regularity.

Finally, the SIGP weak form is the following variational inequality (VI) \cite{KinderlehrerStampacchia1980} which should determine $s\in\mathcal{K}$:
\begin{equation}
F(s)[r - s] \ge \ip{a}{r-s} \quad \text{for all $r \in \mathcal{K}$.}  \label{eq:sigpweakform}
\end{equation}
This VI has a shallow (SIA) analog, for which existence has been proven \cite{JouvetBueler2012}, but in that case the functional is an integral of powers of the thickness $s-b$ and the surface slope $\grad s$, thus locally-computable.  Here, by contrast, the computation of \eqref{eq:sigpfunctional} is non-local and expensive: to evaluate $F(s)[r]$ one must solve Glen-Stokes problem \eqref{eq:glenstokesweak} on the given $z=s(x,y)$ geometry.  In any case one can prove that if $s$ is $C^1$ and solves \eqref{eq:sigpweakform} then it solves strong forms \eqref{eq:strongform} and \eqref{eq:idostrongform} \cite{Bueler2021conservation}.  In particular the strong SKE \eqref{eq:ske} holds on $s>b$.

Definition \eqref{eq:sigpfunctional} is a dual pairing of $F(s)$ and $r$, thus roughly an inner product.  Speaking geometrically, VI \eqref{eq:sigpweakform} says that $s$ is located in $\mathcal{K}$, generically on $\partial\mathcal{K}$, at a place where $F(s)$ points directly into $\mathcal{K}$.  Specifically, \eqref{eq:sigpweakform} says that the ``angle'' between $F(s)$ and an arbitrary vector $r-s$ pointing into $\mathcal{K}$ is at most $90^\circ$.

It is well-known that certain VIs arise as inequality-constrained minimization problems \cite{GraeserKornhuber2009,KinderlehrerStampacchia1980}, but to the best of our knowledge VI \eqref{eq:sigpweakform} does \emph{not} arise in this way.  The general-bed, steady shallow (SIA) model is also not a minimization, except in the flat-bed case \cite{JouvetBueler2012}.

The SIGP weak form \eqref{eq:sigpweakform} is coupled to \eqref{eq:glenstokesweak} through definition \eqref{eq:ido} of the IDO $\Phi$.  The problem therefore has three fundamental nonlinearities:
\renewcommand{\labelenumi}{(\emph{\roman{enumi}})}
\begin{enumerate}
\item the Glen power-law rheology,
\item the inequality constraint, and
\item the nonlinearity of free-surface flow reflected in the nonlinearity of the operator $\Phi$.
\end{enumerate}
Regarding (\emph{ii}), observe that even the classical obstacle problem for the linear Laplacian operator is nonlinear \cite{KinderlehrerStampacchia1980}.  Regarding (\emph{iii}), note that a free-surface flow for Newtonian (linear) rheology produces a nonlinear equation for the flow thickness, e.g.~as expressed in the kinematic wave equation \cite{Ockendonetal2003}.

We will need the (Gateaux) derivative of $F$, and this requires differentiating $\Phi$ as well.  Suppose $s\in \mathcal{K}$, $\eps>0$, and $t \in W_0^{1,\qq}(\Omega)$ is such that $s+\eps t \in \mathcal{K}$.  (In the following, both $s$ and $s+\eps t$ must be admissible glacier surface elevations.)  From \eqref{eq:sigpfunctional}, if $r\in W_{1,\qq}(\Omega)$ then
\begin{equation}
F'(s)[t,r] = \lim_{\eps\to 0^+} \frac{F(s+\eps t)[r] - F(s)[r]}{\eps} = - \int_\Omega \Phi'(s)[t]\, r \,dx dy \label{eq:sigpfunctionalderiv}
\end{equation}
where by definition
\begin{equation}
\Phi'(s)[t] = \lim_{\eps\to 0^+} \frac{\bu|_{s+\eps t} \cdot \bn_{s+\eps t} - \bus \cdot \bn_s}{\eps}. \label{eq:idoderiv}
\end{equation}
In \eqref{eq:idoderiv} note that all quantities are extended to the whole of $\Omega$.

We will assume that $\Phi'(s)[t]$ is in the dual space of $W^{1,\qq}(\Omega)$ so that \eqref{eq:sigpfunctionalderiv} is finite.  Observe that $F'$ is straightforward to compute from $\Phi'$, but further comments on \eqref{eq:idoderiv} are forthcoming.  In fact, $F'(s)[t,r]$ and $\Phi'(s)[t]$ are one-sided directional derivatives in the sense that $s+\eps t$ must be admissible, i.e.~$s+\eps t\ge b$, for all sufficiently-small $\eps>0$, which requires $t\ge 0$ on the (active) set where $s=b$.  However, to simplify our considerations we define an $s$-independent set
\begin{equation}
\mathcal{W}_+ = \{t \in W_0^{1,\qq}(\Omega) \,:\, t(x,y) \ge 0\}. \label{eq:infdefectset}
\end{equation}
It follows that $\Phi'(s)[t]$ is well-defined on inputs $(s,t) \in \mathcal{K} \times \mathcal{W}_+$ and $F'(s)[t,r]$ is well-defined for $(s,t,r) \in \mathcal{K} \times \mathcal{W}_+ \times W^{1,\qq}(\Omega)$.  We will approximate the true derivative $\Phi'(s)[t]$ by the difference quotient in \eqref{eq:idoderiv}, though only in the situation where $t \in \mathcal{W}_+$ has small support.

In fact the meaning of the finite difference in \eqref{eq:idoderiv} depends on the support of $t$:
\begin{equation}
\bu|_{s+\eps t} \cdot \bn_{s+\eps t} - \bus \cdot \bn_s = \begin{cases}
           (\bu|_{s+\eps t} - \bus) \cdot \bn_{s+\eps t} + \eps\, \bus \cdot \left<t_x,t_y,0\right>, & s > b \\
           \bu|_{s+\eps t} \cdot \bn_{s+\eps t}, & s=b, t > 0 \\
           0, & s=b, t = 0.
                 \end{cases} \label{eq:differencecases}
\end{equation}
Thus a necessary condition for differentiability of $\Phi$ is that $\bu|_{s+\eps t} \cdot \bn_{s+\eps t} = o(\eps)$ on the ice-free area $s=b$.  In physical terms this means that the dynamical part of the SKE \eqref{eq:ske} vanishes for small ice masses on bare ground, which is reasonable to suppose if the ice cannot slide.

In the next three sections we will construct an iterative, multilevel finite element solver for SIGP weak form \eqref{eq:sigpweakform}.  When we attempt to approximate the derivative $\Phi'(s)[t]$ by a difference quotient, a significant issue arises because an extra solution of Glen-Stokes problem \eqref{eq:glenstokesweak} is required over the icy domain $\Lambda_{s+\eps t}$, but in Section \ref{sec:smoothers} we propose a less-expensive approximation.

@@

In this section we propose a new multilevel scheme for solving the discrete SIGP weak form \eqref{eq:fesigpweakform} using iterated V-cycles.  The method descends from the multilevel constraint decomposition (MCD) method of \cite{Tai2003} (see also \cite{GraeserKornhuber2009}).  Our nonlinear variant, denoted MCDN, uses a full approximation scheme (FAS) multigrid \cite{Trottenbergetal2001} approach which transfers both the current residual and an approximation of the solution down to coarser levels.

>>>>

\newcommand{\nn}{{\text{n}}}
\newcommand{\pp}{{\text{p}}}
\newcommand{\qq}{{\text{q}}}
\newcommand{\rr}{{\text{r}}}

@@

We model the ice as an incompressible, very-viscous \cite{Acheson1990}, non-Newtonian fluid subject to Glen's shear-thinning flow law \cite{GreveBlatter2009}; see also \cite[Chapter 1]{FowlerNg2021}.  Allowing any Glen exponent $\nn\ge 1$, the equations inside $\Lambda_s$ are
\begin{align}
- \nabla \cdot \tau + \nabla p &= \rhoi \bg &&\text{\emph{stress balance}} \label{eq:forcebalance} \\
\nabla \cdot \bu &= 0 &&\text{\emph{incompressibility}} \label{eq:incompressible} \\
\tau &= B_\nn |D\bu|^{(1/\nn) - 1} D\bu  &&\text{\emph{flow law}} \label{eq:viscflowlaw}
\end{align}
The solution fields here are the velocity $\bu$, pressure $p$, and deviatoric stress $\tau$.  Equations \eqref{eq:forcebalance} and \eqref{eq:viscflowlaw} form the momentum conservation (stress balance) model, while \eqref{eq:incompressible} follows from constant density and mass conservation.

@@

In computations we will use constants $\nn=3$, ice density $\rhoi=910 \,\text{kg}\,\text{m}^{-3}$ \cite{Huybrechtsetal1996}, and gravity $\bg=\left<0,0,-g\right>$, with $g=9.81\,\text{m}\,\text{s}^{-2}$.  The ice hardness $B_\nn$ is a constant because we assume isothermal conditions \cite{GreveBlatter2009}, with the value $B_3=6.8082\times 10^7\,\text{Pa}\,\text{s}^{1/3}$ \cite{Huybrechtsetal1996} used in computations.

For the $\nn=1$ (Newtonian) Stokes equations one would write \eqref{eq:viscflowlaw} as $\tau = 2\nu D\bu$ with viscosity $\nu>0$, but exponents $\nn>1$ imply an effective viscosity function of $|D\bu|$.  This would be singular in the limit of small strain rates, and so, motivated by the actual finite viscosity of glacier ice \cite{GreveBlatter2009}, we define the regularized effective viscosity using $\eps>0$,
\begin{equation}
\nu_\eps = \frac{1}{2} B_\nn \left(|D\bu|^2 + \eps\, D_0^2\right)^{(\pp-2)/2}, \label{eq:regeffvisc}
\end{equation}
where $\pp=(1/\nn)+1$, with $\pp=4/3$ in computations.  The constant $D_0$ defines a strain-rate scale for glacier flow; values $D_0 = 1 \,\text{a}^{-1}$ and $\eps = 10^{-4}$ are used in computations.

@@

We may now state the strong form of the SIGP, as follows, by including all of the above conditions, and also by eliminating $\tau$ from \eqref{eq:forcebalance} and \eqref{eq:viscflowlaw}:
\begin{align}
s - b &\ge 0 && \text{on $\Omega$} \label{eq:strongform} \\
- \bu|_s \cdot \bn_s - a &\ge 0 && \text{\emph{same}} \notag \\
(s - b) (- \bu|_s \cdot \bn_s - a) &= 0 && \text{\emph{same}} \notag \\
- \nabla \cdot \left(2 \nu_\eps\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{on $\Lambda_s$} \notag \\
\nabla \cdot \bu &= 0 && \text{\emph{same}} \notag \\
\bu &= \bzero && \text{on $\Gamma_0$} \notag \\
\left(2 \nu_\eps D\bu - pI\right) \bn &= \bzero && \text{on $\partial \Lambda_s \setminus \Gamma_0$} \notag
\end{align}
(Definition \eqref{eq:regeffvisc} of $\nu_\eps$ is included.)  The first three statements in \eqref{eq:strongform} form the NCP, but this problem is coupled to the boundary value problem formed by the last four statements.

@@

Denote by $W^{k,r}$ the Sobolev space \cite{Evans2010} of functions with $k$ derivatives which are $r$th-power integrable.  Suppose for the moment that $\Lambda \subset \RR^{d+1}$ is a fixed, bounded domain on which we prescribe a Dirchlet condition $\bu=0$ on some positive-measure portion of the boundary, and otherwise suppose the boundary is stress-free (Neumann boundary).  Let $\pp=(1/\nn)+1$ as in \eqref{eq:regeffvisc}, and $\qq=(1-\pp^{-1})^{-1}=\nn+1$ the conjugate exponent; $\pp=4/3$ and $\qq = 4$ if $\nn=3$.  Let $W_0^{1,\pp}(\Lambda)^{d+1}$ be the space of velocity functions, zero along the Dirichlet boundary, and let
\begin{equation}
\mathcal{M}_\Lambda = W_0^{1,\pp}(\Lambda)^{d+1} \times L^\qq(\Lambda)  \label{eq:mixed}
\end{equation}
be the (mixed) space of admissible velocity and pressure pairs.  The Glen-Stokes solution $(\bu,p) \in \mathcal{M}_\Lambda$ should satisfy the weak form
\begin{equation}
F_\Lambda(\bu,p)[\bv,q] = \int_\Lambda 2 \nu_\eps D\bu : D\bv - p \Div\bv - (\Div\bu) q - \rhoi \bg \cdot \bv\,d\bx = 0 \label{eq:glenstokesweak}
\end{equation}
for all $(\bv,q) \in \mathcal{M}_\Lambda$.

@@

However, we must go beyond the problem on predetermined ice geometry to solve the weak form of \eqref{eq:strongform}.  Again suppose $\Omega \subset \RR^d$ is a fixed map-plane domain with $C^1$ boundary, and that the bed elevation $b$ is in $W^{1,\qq}(\Omega)$.  The Sobolev space $W^{1,\qq}(\Omega)$ supports a linear trace operator which defines the symbol $|_{\partial \Omega}$ \cite[Section 5.5]{Evans2010}.  Let
\begin{equation}
\mathcal{K} = \{s \in W^{1,\qq}(\Omega) \,:\, s \ge b \, \text{ and } \, s\big|_{\partial\Omega} = b\big|_{\partial\Omega}\}  \label{eq:Kconstraintset}
\end{equation}
be the closed and convex set of admissible surface elevations of an isolated glacier or ice sheet.  From now on we also assume that $\nn>1$.  Because $\qq = \nn+1 > d$, each $s \in W^{1,\qq}(\Omega)$ is continuous \cite[Morrey's inequality, section 5.6.2]{Evans2010}, and so there are no cliffs in the $s$ geometry.  (Though it is not clear that the correct Sobolev space is identified in \eqref{eq:Kconstraintset}, the constraint is surely correct.  Existence for the simpler SIA model finds a power of the ice thickness, $u=(s-b)^{2q/(q-1)}$, in $\mathcal{K} = \{v \ge 0\} \subset W^{1,\qq}(\Omega)$ \cite{JouvetBueler2012}.)

Recalling definition \eqref{eq:lambdas}, we further suppose that
\begin{equation}
s\in \mathcal{K} \text{ defines an icy domain } \Lambda_s \text{ which has a $C^1$ upper surface.} \label{eq:quixotic}
\end{equation}
Assumption \eqref{eq:quixotic} would follow from a sufficiently-strong regularity result for the theory of the current paper, which we cannot offer.  In the SIA model the surface $s$ solves an elliptic equation, so it is smooth on $\{s>b\} \subset \Omega$ if $b$ is itself smooth \cite{JouvetBueler2012}.

We now define the ice dynamics operator (IDO), a map $\Phi$ which uses the trace of the velocity solution to the Glen-Stokes problem above (Figure \ref{fig:idoaction}).  For each surface elevation function $s \in \mathcal{K}$ the output $\Phi(s)$ is the negative normal component of the (unique) velocity solution $\bu$ of \eqref{eq:glenstokesweak} over the upper surface of $\Lambda_s$, but extended by zero to all of $\Omega$:
\begin{equation}
\Phi(s) = \begin{cases} - \bu|_s \cdot \bn_s, & s > b, \\
                        0, & \text{otherwise}. \end{cases} \label{eq:ido}
\end{equation}
Here $\bu|_s$ denotes a trace operator on $W^{1,\pp}(\Lambda_s)^{d+1}$; compare \eqref{eq:mixed}.  Note that $\Phi(s)$ is a scalar function on $\Omega$ which may jump or have other singularities at the glacier margin, but we will only need $\Phi(s)$ to be in the dual of $W^{1,\qq}(\Omega)$.

\begin{figure}[t]
\begin{center}
\includegraphics[width=\textwidth]{genfigs/idoaction.pdf}
\end{center}
\caption{The ice dynamics operator $\Phi$ maps the surface elevation $s$ (dashed) to the normal ice surface motion $\Phi(s)=- \bu|_s \cdot \bn_s$, equation \eqref{eq:ido}.  Both the input and output of $\Phi$ are defined on all of $\Omega$.}
\label{fig:idoaction}
\end{figure}

We believe that formula \eqref{eq:ido} defines a nonlinear operator $\Phi:\mathcal{K} \subset W^{1,\qq}(\Omega) \to (W^{1,\qq}(\Omega))^* = W^{-1,\pp}(\Omega)$.  This operator is \emph{expensive}; evaluating $\Phi(s)$ requires solving Glen-Stokes problem \eqref{eq:glenstokesweak} on $\Lambda_s \subset \RR^{d+1}$ defined in \eqref{eq:lambdas}.  Furthermore the operator is clearly \emph{nonlocal} because the solution $\bu$ is influenced everywhere by changes in the icy domain geometry.  In particular, if $\tilde s=s + r$ defines $\Lambda_{\tilde s} \supset \Lambda_s$, for $r\in C_c^\infty(\Omega)$ with $r\ge 0$ supported inside $\{s>b\}$, then the Stokes solutions are generically different everywhere both solutions are defined: $\tilde\bu - \bu \ne 0$ on all of $\Lambda_s$, including the trace on the upper surface.  Thus an FE discretization of $\Phi(s)$ will have a dense Jacobian matrix

By contrast, in the simpler SIA model the operator is a nonlinear, second-order differential operator,
\begin{equation}
\Phi_{\text{SIA}}(s) = - \frac{\gamma}{\qq} (s-b)^{\qq} |\grad_2 s|^{\qq} - \grad_2 \cdot\left[\frac{\gamma}{\qq+1} (s-b)^{\qq+1} |\grad_2 s|^{\qq-2} \grad_2 s\right], \label{eq:phisia}
\end{equation}
for $\gamma = 2(\rhoi g)^{\nn} (B_\nn)^{-\nn} > 0$ and $\grad_2 = (\partial_x,\partial_y)$.  Thus an FE discretization of $\Phi_{\text{SIA}}(s)$ has sparse Jacobian.  However, $\Phi_{\text{SIA}} \approx \Phi$ for shallow glaciers to the degree that the lubrication approximation \cite{Acheson1990} is valid.

FIXME THREE IDO REASONS TO CARE: EXPRESS STRONG AND WEAK FORM, DIAGONAL OF IDO UNKNOWN BUT POSITIVE? IN SMOOTHER, SPECTRUM OF IDO DETERMINES STABILITY OF TIME-STEPPING

Now we may re-write \eqref{eq:strongform} more simply as an NCP with a hidden dynamical problem:
\begin{align}
s - b &\ge 0  \label{eq:idostrongform} \\
\Phi(s) - a &\ge 0 \notag \\
(s - b) (\Phi(s) - a) &= 0 \notag
\end{align}
Here the statements hold on all of $\Omega$ and the roles of the data $a,b$ are evident.  The solution is the ice surface $z=s(x,y)$ while the variables $\bu,p$ are hidden within the evaluation of $\Phi(s)$.

For $s \in \mathcal{K}$ and $r \in W^{1,\qq}(\Omega)$ we next define a ``residual'' functional which is simply the dual of $\Phi$:
\begin{equation}
F(s)[r] = \ip{\Phi(s)}{r} = \int_\Omega \Phi(s)\, r \,dx dy. \label{eq:sigpfunctional}
\end{equation}
One computes $F(s)[r]$ by defining $\Lambda_s$ as in \eqref{eq:lambdas}, solving \eqref{eq:glenstokesweak}, evaluating the surface trace of the velocity to compute $\Phi(s)$, and then integrating the result against (i.e.~dual pairing with) the test function $r$.

The SIGP weak form is the following variational inequality (VI) \cite{KinderlehrerStampacchia1980} for $s\in\mathcal{K}$:
\begin{equation}
F(s)[r - s] \ge \ip{a}{r-s} \quad \text{for all $r \in \mathcal{K}$.}  \label{eq:sigpweakform}
\end{equation}
One can prove that if $s$ is $C^1$ and solves \eqref{eq:sigpweakform} then it solves strong forms \eqref{eq:strongform} and \eqref{eq:idostrongform} \cite{Bueler2021conservation}.  In particular the strong SKE \eqref{eq:ske} holds on $s>b$.

Speaking geometrically, \eqref{eq:sigpweakform} says that $s$ is located in $\mathcal{K}$, generically on $\partial\mathcal{K}$, at a place where $\Phi(s)-a$ points directly into $\mathcal{K}$.  The VI says that the ``angle'' between $\Phi(s)-a$ and an arbitrary vector $r-s$ pointing into $\mathcal{K}$ is at most $90^\circ$.  Also, it is well-known that certain VIs arise as inequality-constrained minimization problems \cite{GraeserKornhuber2009,KinderlehrerStampacchia1980}, but to the best of our knowledge VI \eqref{eq:sigpweakform} does \emph{not} arise in this way.

Existence has been proven for the SIA analog of VI \eqref{eq:sigpweakform} \cite{JouvetBueler2012}, but, as noted above, in that case the residual is locally-computable.  The general-bed, steady shallow (SIA) model is also not a minimization, except in the flat-bed case \cite{JouvetBueler2012}.

The SIGP weak form \eqref{eq:sigpweakform} is coupled to \eqref{eq:glenstokesweak} through definition \eqref{eq:ido} of the IDO $\Phi$.  The VI problem therefore has three fundamental nonlinearities:
\renewcommand{\labelenumi}{(\emph{\roman{enumi}})}
\begin{enumerate}
\item the Glen power-law rheology,
\item the inequality constraint, and
\item the nonlinearity of free-surface flow as a function of the surface.
\end{enumerate}
Regarding (\emph{ii}), observe that the solution of the classical obstacle problem for the linear Laplacian operator is a nonlinear function of the data \cite{KinderlehrerStampacchia1980}.  Regarding (\emph{iii}), note that a free-surface flow for Newtonian (linear) rheology produces a nonlinear equation for the flow thickness, e.g.~as expressed in the kinematic wave equation \cite{Ockendonetal2003}.

We will need the (Gateaux) derivative of $F$, thus $\Phi$.  Suppose $s\in \mathcal{K}$, $\eps>0$, and $t \in W_0^{1,\qq}(\Omega)$ is such that $s+\eps t \in \mathcal{K}$.  From \eqref{eq:sigpfunctional}, if $r\in W^{1,\qq}(\Omega)$ then $F'(s)[t,r] = \int_\Omega \Phi'(s)[t]\, r$ where by definition
\begin{equation}
\Phi'(s)[t] = - \lim_{\eps\to 0^+} \frac{\bu|_{s+\eps t} \cdot \bn_{s+\eps t} - \bus \cdot \bn_s}{\eps}, \label{eq:idoderiv}
\end{equation}
extended by zero to $\Omega$.  However, $\Phi'(s)[t]$ is actually a one-sided directional derivative on a portion of $\Omega$.  That is, $s+\eps t\ge b$ is required for all sufficiently-small $\eps>0$, which requires $t\ge 0$ on the (active) set where $s=b$.  To simplify we define an $s$-independent set
\begin{equation}
\mathcal{W}_+ = \{t \in W_0^{1,\qq}(\Omega) \,:\, t(x,y) \ge 0\}. \label{eq:infdefectset}
\end{equation}
It follows that $\Phi'(s)[t]$ is well-defined on inputs $(s,t) \in \mathcal{K} \times \mathcal{W}_+$, thus $F'(s)[t,r]$ is well-defined for $(s,t,r) \in \mathcal{K} \times \mathcal{W}_+ \times W^{1,\qq}(\Omega)$.

The meaning of the difference in \eqref{eq:idoderiv} depends on the support of $t$:
\begin{equation}
\bu|_{s+\eps t} \cdot \bn_{s+\eps t} - \bus \cdot \bn_s = \begin{cases}
           (\bu|_{s+\eps t} - \bus) \cdot \bn_{s+\eps t} + \eps\, \bus \cdot \left<t_x,t_y,0\right>, & s > b \\
           \bu|_{s+\eps t} \cdot \bn_{s+\eps t}, & s=b, t > 0 \\
           0, & s=b, t = 0.
                 \end{cases} \label{eq:differencecases}
\end{equation}
Thus a necessary condition for differentiability of $\Phi$ is that $\bu|_{s+\eps t} \cdot \bn_{s+\eps t} = o(\eps)$ on the ice-free area $s=b$.  In physical terms, the ice motion part of the SKE \eqref{eq:ske} must vanish for small ice masses on bare ground, a reasonable supposition if the ice cannot slide.

In the next three sections we will construct an iterative, multilevel finite element solver for SIGP weak form \eqref{eq:sigpweakform}.  When we attempt to approximate the derivative $\Phi'(s)[t]$ by a difference quotient, significant performance and stability issues arise.  These are addressed in Section \ref{sec:smoothers}.

@@

In this section we propose a new multilevel scheme for solving the discrete SIGP weak form \eqref{eq:fesigpweakform} using iterated V-cycles.  The method descends from the multilevel constraint decomposition (MCD) method of \cite{Tai2003}, see also \cite{GraeserKornhuber2009}, but our ``MCDN'' nonlinear variant uses a full approximation scheme (FAS) multigrid \cite{Trottenbergetal2001} approach which transfers both the current residual and an approximation of the solution down to coarser levels.
